---
title: "Centering in Mixed Models"
description: |
  Yadda Yadda
author:
  - name: Michael Clark
    url: https://m-clark.github.io
date: '2020-05-14'
preview: ../../img/covid_preview.gif  
output:
  distill::distill_article:
    self_contained: false
    toc: true
    css: [../../styles.css, ../../css/misc.css]
    code_folding: hide
bibliography: covid.bib
draft: true
tags: [mixed models, group centering, grand mean centering]
categories:
  - visualization
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = NA,
  R.options = list(width = 120),
  cache.rebuild = FALSE,
  cache = FALSE,
  fig.align = 'center',
  fig.asp = .7,
  dev = 'svglite',
  dev.args = list(bg = 'transparent')
)

library(tidyverse)
library(kableExtra)
library(scico)

kable_df <- function(..., digits=3) {
  kable(..., digits=digits) %>% 
    kable_styling(full_width = F)
}

rnd = tidyext::rnd
```

<aside>First posted 2020-05-14.</aside>

## Introduction

Extending Bell's articles

## Data Example


## Simulation

### Setup

Model:
- y continuous vs. binary  (binary only after continuous is complete)
- level 1: x with mildly correlated q
- level 2: avg x with std coef 0, .1, .25; some other mildly correlated cluster level covariate


Considerations: 
- n-per-group  5, 20, 100,
- n-groups  doesn't appear to matter, but maybe do very large 500, 1000, since no one else appears to
- icc: 50% vs. 10%

```{r}
ng = 500
npg = 5


x = rnorm(1000)
```


### Results