---
title: "Marginal Effects for Mixed Models with brms"
author:
  - name: Michael Clark
    url: https://m-clark.github.io
date: '`r format(Sys.Date(), "%B %d, %Y")`'
preview: <i class="fas fa-percent fa-5x" style = 'color:#ff5500'></i>
output:
  distill::distill_article:
    self_contained: false
    toc: true
    css: ../../styles.css
draft: true
tags: [regression, marginal effects, brms, mixed models, bayesian]
categories:
  - regression
  - mixed models
  - brms
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE, 
  comment = NA,
  R.options = list(width = 120),
  cache.rebuild = FALSE,
  cache = TRUE, 
  fig.align='center', 
  dev = 'svg', 
  dev.args=list(bg = 'transparent')
)

library(tidyverse)
library(tidyext)
library(broom)
library(kableExtra)
library(visibly)

kable_df <- function(..., digits = 3) {
  kable(..., digits = digits) %>%
    kable_styling(full_width = F)
}
```

https://github.com/paul-buerkner/brms/issues/552
https://github.com/jgabry/posterior/issues/39

## Linear Models

```{r}
margex = haven::read_dta('data/margex.dta')

m <- margex %>% 
  data.frame() %>% 
  mutate_at(vars(sex, group), as_factor)

head(m)
```

### Marginal Means

#### Categorical

```{r}
library(brms)
mod_cat <- brm(y ~ group, data = m, cores = 4)

summary(mod_cat)
```

```{r}
mod_cat_me = marginal_effects(mod_cat)

mod_cat_me$group %>% 
  select(group, estimate__:upper__) %>% 
  kable_df()

mod_cat_me
```

#### Continuous



```{r}
mod_cont <- brm(y ~ sex + age + distance, data = m)

summary(mod_cont)
```

At present we can only get marginal means as predictions at key values from brms. 

```{r}
cd = make_conditions(mod_cont, vars = c('sex', 'age', 'distance'))
cd
mod_cont_me = fitted(mod_cont, newdata = cd)
mod_cont_me


cd = data.frame(
  expand.grid(
    sex = c('male', 'female'),
    age = c(30, 40)
  ),
  distance = mean(m$distance)
)
mod_cont_me = fitted(mod_cont, newdata = cd)
data.frame(cd, mod_cont_me)
```

But since this is a basic linear model, we can use emmeans to get the average expected value at age 30 and 40.

```{r}
emmeans::emmeans(mod_cont, specs = ~ age, at = list(age = c(30, 40)))
emmeans::emmeans(mod_cont, ~ age + distance, at = list(age = c(30, 40), distance = c(10,100)))
emmeans::emmeans(mod_cont, ~ age + sex, at = list(age = c(30, 40)))  # essentially the same as when dist was held at mean
```


#### Interactions 

##### Categorical by Categorical

```{r}
mod_cat_by_cat <- brm(y ~ sex*group, data = m)

summary(mod_cat_by_cat)
```

```{r}
mod_cat_by_cat_me = marginal_effects(mod_cat_by_cat, effects = 'sex:group')

mod_cat_by_cat_me$`sex:group` %>% 
  select(sex, group, estimate__:upper__) %>% 
  kable_df()

mod_cat_by_cat_me
```



##### Categorical by Continuous

```{r}
mod_cat_by_cont <- brm(y ~ sex*age, data = m)

summary(mod_cat_by_cont)
```

```{r}
mod_cat_by_cont_me = marginal_effects(mod_cat_by_cont, effects = 'sex:age', int_conditions = list(age = c(30, 40)))

mod_cat_by_cont_me$`sex:age` %>% 
  select(sex, age, estimate__:upper__) %>% 
  arrange(desc(age)) %>% 
  kable_df()

mod_cat_by_cont_me
```


### Marginal Slopes

```{r}
mod_cont_me = marginal_effects(mod_cont, effects = 'age')

mod_cont_me$age %>% 
  select(age, estimate__:upper__) %>% 
  head() %>% 
  kable_df()

mod_cont_me
```


#### Pairwise comparisons

### Continuous

### Interactions

```{r}

noiris::instructor_evaluations 
```


## Generalized Linear Models

### Categorical

### Continuous

### Interactions

```{r}
fit_zinb1 <- brm(count ~ persons + child + camper, 
                 data = zinb,
                 family = zero_inflated_poisson("log"))
```

