---
title: "Group differences on the unseen"
description: |
  Comparing means and intercepts of latent variables
author:
  - name: Michael Clark
    url: https://m-clark.github.io
date: '`r format(Sys.Date(), "%B %d, %Y")`'
preview: ../../img/198R.png   # apparently no way to change the size displayed via css (ignored) or file (stretched)
output:
  distill::distill_article:
    self_contained: false
    toc: true
    css: ../../test.css
draft: true
tags: [R, factor analysis, growth curve, structural equation modeling, SEM, intercepts, means]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo=T, 
  message = F, 
  warning = F, 
  comment = NA,
  R.options=list(width=120), 
  cache.rebuild=F, 
  cache=FALSE,
  fig.align='center', 
  dev = 'svg', 
  dev.args=list(bg = 'transparent')
)

library(tidyverse); library(broom); library(kableExtra); library(visibly)

kable_df <- function(..., digits=3) {
  kable(..., digits=digits) %>% 
    kable_styling(full_width = F)
}

rnd = function(x, digits = 3) arm::fround(x, digits = digits)
```



## Introduction

In some cases we are interested in looking at group differences in latent variables.  For example, social scientists are interested in race and sex differences on psychological measures, or educational scientists might want to create exams in different languages. There are a variety of ways in to assess group differences across latent structure, such as anxiety or verbal ability, and this post provides a demo using lavaan.  

My motivation for doing this is that it comes up from time to time in consulting, and I wanted a quick reminder for the syntax to refer back to.  As a starting point though, you can find some demonstration on the [lavaan website](https://lavaan.ugent.be).  For more on factor analysis, structural equation modeling, and more, see [my document](https://m-clark.github.io/sem/).

## Multiple group analysis

A common way to assess group differences is via multiple group analysis, which amounts to doing separate structural equation models of some kind across the groups of interest.  We will use a classic data set to demonstrate the approach. From the help file:

> The Holzinger and Swineford (1939) dataset consists of mental ability test scores of seventh- and eighth-grade children from two different schools (Pasteur and Grant-White). In the original dataset, there are scores for 26 tests. However, a smaller subset with 9 variables is more widely used in the literature...

```{r data}
library(lavaan)
data(HolzingerSwineford1939)
```

The basic model is a factor analysis with three latent variables, with items for visual-spatial ability (`x1-x3`), verbal comprehension (`x4-x6`), and so-called 'speed' tests (`x7-x9`), e.g. for addition and counting, which might be thought of general cognitive processing.

With lavaan, we specify the model for three factor (or latent variables).  After that, a simple group argument will allow the multigroup analysis, providing the factor analysis for both school groups.

```{r standard_multigroup}
library(tidyverse)
library(lavaan)

hs_model_baseline <- ' 
  visual =~ x1 + x2 + x3
  verbal =~ x4 + x5 + x6
  speed  =~ x7 + x8 + x9 
'

fit_baseline <- cfa(
  hs_model_baseline, 
  data = HolzingerSwineford1939, 
  group = "school"
)

summary(fit_baseline)  
```

So we're left with visual inspection to note whether there are general differences on latent variables among the groups. This is all well and good, but perhaps we want a more statistical approach.  Say our question specifically concerns a mean difference between schools on the visual latent variable.  How do we go about it?

Note that at this point the intercepts for the latent variables are zero.  They have to be for the model to be identified, much in the same way that at least one factor loading (the first by default) has to be fixed to one.  We only have so much information to estimate so many parameters.  Now let's see how we might go about changing things to get a better understanding of group differences on the latent variables.

## Latent variable intercepts

```{r baseline_2, eval=FALSE, echo=F}
fit_baseline_2 <- cfa(
  hs_model_baseline, 
  data = HolzingerSwineford1939, 
  group = "school",
  group.equal = c('loadings', 'intercepts')
)

summary(fit_baseline_2) 
```


To get around this limitation, we could try and fix some parameters, thereby freeing the intercepts to be estimated.  For example, if we fix the mean of one of the observed variables to be zero instead, we would be able to estimate the intercept for the latent variable. In the following we'll do this for the visuo-spatial ability construct, which will be our point of focus for group differences going forward.  

In the model we we also identify a new parameter, which will be the differences in these latent variable intercpets, simply called `diff`.  I have omitted some output for brevity of space. 

```{r hs_model_1}
hs_model_1 <- ' 
  visual =~ x1 + x2 + x3
  verbal =~ x4 + x5 + x6
  speed  =~ x7 + x8 + x9 
  
  
  # intercepts: in order to have an identified model, you would have to fix the
  # intercepts of observed to 0, 1 represents the intercept, 0* fixes it to be 0
  x1 ~ 0*1   

  # intercept for Pasteur and Grant-White schools
  visual ~  c(int_p, int_gw)*1    
   
  # comparisons
  diff := int_p - int_gw
'

fit_1 <- cfa(hs_model_1, 
             data = HolzingerSwineford1939, 
             group = "school",
             meanstructure = T)
summary(fit_1, header=F, nd=2)
```


For clearer presentation, we'll look at a table of the specific parameter estimates.

```{r hs_model_1_table, echo=FALSE}
coef_1 = broom::tidy(fit_1) %>% 
  filter(label != '')

coef_1 %>% kable_df()
```


The above shows the schools not to be much different from one another on the visual-spatial ability latent variable.  But compare this result to the intercepts for `x1` in our baseline model.  This model would would be identical to comparing the intercepts on whichever observed variable you previously fixed to zero.  Much like we must scale the latent variable to that of one of the observed variables by fixing the loading to be 1, we essentially come to the same type of issue by fixing its mean to be on that of the observed variable.

To make this more explicit, we'll label the `x1` intercepts in our baseline model and look at their difference.  I won't show the model out put and simply focus on the parameter table instead.

```{r hs_model_2}
hs_model_2 <- ' 
  visual =~ x1 + x2 + x3
  verbal =~ x4 + x5 + x6
  speed  =~ x7 + x8 + x9 
  
  x1 ~ c(a, b)*1   
   
  # comparisons
   diff := a - b
'

fit_2 <- cfa(hs_model_2, 
           data = HolzingerSwineford1939, 
           group = "school",
           meanstructure = T)

# summary(fit_2)
```

```{r, hs_model_2_table, echo=FALSE}
coef_2 = broom::tidy(fit_2) %>% 
  filter(label != '')

coef_2 %>% kable_df()
```


Same difference.


## Observed variable group differences

The following approach is not the same model, but would also provide the same result. In this case, each observed variable is affected by the school grouping, and the path coefficient for `x1` is the same difference in means as before.


```{r hs_model_3}
hs_model_3 <- ' 
  visual =~ x1 + x2 + x3
  verbal =~ x4 + x5 + x6
  speed  =~ x7 + x8 + x9 
  
  x1 ~ diff*school 
  x2 + x3 + x4 + x5 + x6 +  x7 + x8 + x9 ~ school
'

fit_3 <- cfa(hs_model_3, 
             data = HolzingerSwineford1939,
             meanstructure = T)

# summary(fit_3)
```

A comparison of all three shows the same results, but that the third model has fewer parameters, as the loadings and latent variable variances are not changing across groups.

```{r compare_123_data, echo=FALSE}
coef_3 = broom::tidy(fit_3)

list(
  fit_1 = coef_1 %>% filter(label == 'diff'),
  fit_2 = coef_2 %>% filter(label == 'diff'), 
  fit_3 = coef_3 %>% filter(label == 'diff')
) %>% 
  bind_rows(.id = 'model') %>% 
  select(model, estimate, std.error, conf.low, conf.high) %>% 
  kable_df()

model_list = list(fit_1=fit_1, fit_2=fit_2, fit_3=fit_3)
```


```{r compare_123_Npar, echo=FALSE}
model_list %>% 
  map_df(function(x) length(coef(x))) %>% 
  kable_df(caption = 'Model N parameters')
```


```{r compare_123_AIC, echo=FALSE}
model_list %>% 
  map_df(AIC) %>% 
  kable_df(caption = 'Model AIC')
```

## Structural model

In the models I see, people would more commonly address such a theoretical question without a multigroup approach, simply regressing the latent variable of interest on the group factor.  We can do that here.

```{r hs_model_4a}
# standard cfa with school predicting visual
hs_model_4a <- ' 
  visual =~ x1 + x2 + x3
  verbal =~ x4 + x5 + x6
  speed  =~ x7 + x8 + x9 
  
  visual ~ diff*school
  
  visual ~~ speed + verbal  # lavaan will not estimate this by default
'

fit_4a = sem(hs_model_4a, data=HolzingerSwineford1939, meanstructure=T)
summary(fit_4a)
```

At first blush, it would seem we are not getting the same result, as the mean difference is now estimated to be `r rnd(fit_4a@ParTable$est[fit_4a@ParTable$label=='diff'])`. Our difference is notably larger and significant.  We can also see that tghe fit is different based on AIC and the number of parameters.

```{r compare_1234a, echo=FALSE}
model_list = mget(ls(pattern = '^fit'))

model_list %>% 
  map_df(AIC)
model_list %>% map_df(function(x) length(coef(x)))
```

However, we can recover the previous multigroup results by regressing the other observed variables on school as well.

```{r hs_model_4b}
hs_model_4b <- ' 
  visual =~ x1 + x2 + x3
  verbal =~ x4 + x5 + x6
  speed  =~ x7 + x8 + x9 
  
  visual ~ diff*school
  
  x2 + x3 + x4 + x5 + x6 +  x7 + x8 + x9 ~ school
'

fit_4b = sem(hs_model_4b, data=HolzingerSwineford1939, meanstructure=T)
```

```{r compare_1234b, echo=FALSE}
model_list = mget(ls(pattern = '^fit'))

coef_4b = broom::tidy(fit_4b)

model_list %>% 
  map_df(AIC) %>% 
  kable_df()
model_list %>% 
  map_df(function(x) length(coef(x))) %>% 
  kable_df()

bind_rows(
  coef_1 %>% filter(label == 'diff'), 
  coef_2 %>% filter(label == 'diff'), 
  coef_3 %>% filter(label == 'diff'),
  coef_4b %>% filter(label == 'diff'), 
  .id = 'model'
) %>% 
  mutate(model = c(1:3, '4b')) %>% 
  select(model, estimate, std.error, conf.low, conf.high)
```

Going back to the first structural model `hs_model_4a`, it might be interesting to some to see that it the group difference *still* regards a difference on the observed `x1` observed variable.  We can see this more clearly if we set the `x1` loading to be estimated rather than fixed at one, then use the product of coefficients approach (a la [mediation]()) to estimate the group difference.

```{r hs_model_4c}
hs_model_4c <- ' 
  visual =~ x2 + a*x1 + x3    # estimate x1 loading vs. scaling by it
  verbal =~ x4 + x5 + x6
  speed  =~ x7 + x8 + x9 

  visual ~ b*school
  
  visual ~~ verbal + speed
  
  diff := a*b    # "indirect" effect of school on x1
'

# same as fit_4a
fit_4c = cfa(hs_model_4c, data = HolzingerSwineford1939, meanstructure=T)
```

```{r hs_model_4c_table, echo=FALSE}
coef_4c = broom::tidy(fit_4c) %>% filter(label=='diff') 
coef_4a = broom::tidy(fit_4a) %>% filter(label=='diff') 

bind_rows(coef_4a, coef_4c) %>% 
  mutate(model = c('4a', '4c')) %>% 
  select(model, label:conf.high) %>% 
  kable_df()
```

And what is this value of `r rnd(coef_4c$estimate)`?  We see it as the group difference on the latent construct rather than simply being a mean difference on `x1`.  However, in `fit_4a` it is estimated on the metric of `x1`, which had it's loading fixed to 1. We will see another interpretation later.

## More structural models

The following is equivalent to the result one would get from `group.equal = c('loadings', 'intercepts')`, but to make things more clear, I show the explicity syntax.  The first group would have latent variable means at zero, while the second group would be allowed to vary.  This is more or less what is desired if we want to know a group difference on the latent structure. The first group mean is arbitrarily set to zero, so the estimated intercept for the second group tells us the relative difference.

```{r hs_model_4d}
hs_model_4d <- ' 
  # make loadings equal across groups
  
  visual =~ c(1, 1)*x1 + c(v_x2, v_x2)*x2 + c(v_x3, v_x3)*x3
  verbal =~ c(1, 1)*x4 + c(v_x5, v_x5)*x5 + c(v_x6, v_x6)*x6
  speed  =~ c(1, 1)*x7 + c(v_x8, v_x8)*x8 + c(v_x9, v_x9)*x9 
  
  # make intercepts equal across groups
  
  x1 ~ c(0, 0) * 1
  x2 ~ c(int_x2, int_x2) * 1
  x3 ~ c(int_x3, int_x3) * 1
  x4 ~ c(0, 0) * 1
  x5 ~ c(int_x5, int_x5) * 1
  x6 ~ c(int_x6, int_x6) * 1
  x7 ~ c(0, 0) * 1
  x8 ~ c(int_x8, int_x8) * 1
  x9 ~ c(int_x9, int_x9) * 1
  
  # make covariances equal across groups
  
  # visual ~~ c(cov_vv, cov_vv) * verbal + c(cov_visp, cov_visp) * speed
  # verbal ~~ c(cov_vesp, cov_vesp) * speed
  
  # make variances equal
  
  # visual ~~ c(vvar, vvar) * visual
  # verbal ~~ c(tvar, tvar) * verbal
  # speed  ~~ c(svar, svar) * speed
  
  # x1 ~~ c(x1var, x1var) * x1
  # x2 ~~ c(x2var, x2var) * x2
  # x3 ~~ c(x3var, x3var) * x3
  # x4 ~~ c(x4var, x4var) * x4
  # x5 ~~ c(x5var, x5var) * x5
  # x6 ~~ c(x6var, x6var) * x6
  # x7 ~~ c(x7var, x7var) * x7
  # x8 ~~ c(x8var, x8var) * x8
  # x9 ~~ c(x9var, x9var) * x9
  
  
  visual ~ c(vis_int_p, vis_int_gw)*1
  verbal ~ c(verb_int_p, verb_int_gw)*1
  speed  ~ c(speed_int_p, speed_int_gw)*1
  
  
   
  # comparisons
   diff := vis_int_p - vis_int_gw
'

fit_4d  = sem(hs_model_4d, 
                data=HolzingerSwineford1939, 
                group = 'school',
                meanstructure=T)

summary(fit_4d, header=F, nd=2)
```

The fit is now estimated to be `r rnd(fit_4d@ParTable$est[fit_4d@ParTable$label=='diff'])`, as we are now taking into account the intercorrelations of the latent variables.  However, there is a more simple and obvious way to do this model.  We simply regress all latent variables on school.

```{r hs_model_4e, echo=T}
hs_model_4e <- ' 
  visual =~ x1 + x2 + x3    # estimate x1 loading vs. scaling by it
  verbal =~ x4 + x5 + x6
  speed  =~ x7 + x8 + x9 

  visual ~ diff*school
  
  verbal + speed ~ school
'

fit_4e = cfa(hs_model_4e, data = HolzingerSwineford1939, meanstructure=T)
# summary(fit_4e)
```

```{r hs_model_4e_table, echo=FALSE}
coef_4d = broom::tidy(fit_4d) %>% filter(label=='diff')
coef_4e = broom::tidy(fit_4e) %>% filter(label=='diff')

bind_rows(
  coef_1 %>% filter(label == 'diff'), 
  coef_2 %>% filter(label == 'diff'), 
  coef_3 %>% filter(label == 'diff'),
  coef_4b %>% filter(label == 'diff'), 
  coef_4a %>% filter(label == 'diff'), 
  coef_4c %>% filter(label == 'diff'), 
  coef_4d %>% filter(label == 'diff'), 
  coef_4e %>% filter(label == 'diff'), 
  .id = 'model'
) %>% 
  mutate(model = c(1:3, '4b', '4a', '4c','4d', '4e')) %>% 
  select(model, estimate, std.error, conf.low, conf.high) %>% 
  kable_df() %>% 
  pack_rows(group_label = 'Observed value differences', 1, 4) %>% 
  pack_rows(group_label = 'Latent variable differences', 5, 8)
  
```

Note that the primary differences can be summarized as follows:

- 1:3 and 4b: These models are focusing on observed variable differences, specifically on X1.
- 4a and 4c: These models are latent variable differences on the visual factor, but do not control for indirect (or *backdoor*) effect school has on visual *through* speed and verbal.  In that light, we might consider this the *total effect* of school on the visual factor.  Had we regressed the verbal and speed factors on school also, thereby decomposing the total effect into those different paths, we'd get the same result for the school difference on the visual factor.
- 4d and 4e:  This is generally what we want.  A simple group difference on a latent variable(s) with other parameters assumed (relatively) equal across groups.

## Sum/Factor score

What would happen if we look at the structural/regression model with the estimated latent variable scores[^fsr]?  How about we go even simpler, by not even running an SEM and simply using sum scores?  Let's see what results. 

We'll start with the estimated factor scores. 

```{r lv_score_1}
hs_model_5 <- ' 
  visual =~ x1 + x2 + x3
  verbal =~ x4 + x5 + x6
  speed  =~ x7 + x8 + x9
'

fit_5 = cfa(hs_model_5, data = HolzingerSwineford1939, meanstructure=T)

HolzingerSwineford1939 = HolzingerSwineford1939 %>% 
  mutate(
    visual = lavPredict(fit_5)[,'visual'],
    verbal = lavPredict(fit_5)[,'verbal'],
    speed  = lavPredict(fit_5)[,'speed']
  )

lm_1 = lm(visual ~ school + verbal + speed, data = HolzingerSwineford1939)
coef(lm_1)  
```

The estimated coefficient is pretty close to that estimated by the SEM.  Interestingly, if we fix the loadings to be constant, we recover the initial multigroup estimates.

```{r lv_score_2}
hs_model_5b <- ' 
  visual =~ x1 + l1*x2 + l1*x3
  verbal =~ x4 + l2*x5 + l2*x6
  speed  =~ x7 + l3*x8 + l3*x9 
'

fit_5b = cfa(hs_model_5b, data = HolzingerSwineford1939, meanstructure=T)

HolzingerSwineford1939 = HolzingerSwineford1939 %>% 
  mutate(visual = lavPredict(fit_5b)[,'visual'],
         verbal = lavPredict(fit_5b)[,'verbal'],
         speed  = lavPredict(fit_5b)[,'speed'])
lm_2 = lm(visual ~ school, data = HolzingerSwineford1939)
coef(lm_2)  # same as x1 diffs

lm_3 = lm(x1 ~ school, data = HolzingerSwineford1939)
coef(lm_3)
```

```{r lm_misc, eval=FALSE, echo=FALSE}
lm_2 = lm(resid(lm(visual ~ verbal + speed, data = HolzingerSwineford1939)) ~ school, data = HolzingerSwineford1939)
coef(lm_2)  

lm_3 = lm(visual ~ school + verbal + speed, data = HolzingerSwineford1939)
coef(lm_3)  
```


```{r fsr, eval=FALSE, echo=FALSE}
hs_model_fsr <- ' 
  visual =~ x1 + x2 + x3
  verbal =~ x4 + x5 + x6
  speed  =~ x7 + x8 + x9 
  
  visual ~ school
'

# fsr function now hidden and evidently broken? even naive method doesn't work;
# tried all methods for scores and approach
# ttest1b = lavaan:::fsr(hs_model_fsr,
#                        data = HolzingerSwineford1939)
# coef(ttest1b)

```

Now lets do a sum score. It may not be obvious, but a sum score can be seen as assuming a latent variable model where there is only a single construct and loadings and variances are equal for each item.  As such, it is a natural substitue for a latent variable if we don't want to use SEM.


```{r sum_score}
HolzingerSwineford1939 = HolzingerSwineford1939 %>% 
  rowwise() %>% 
  mutate(visual_sum = sum(x1, x2, x3),
         verbal_sum = sum(x4, x5, x6),
         speed_sum  = sum(x7, x8, x9)) %>% 
  ungroup()
lm_sum = lm(visual_sum ~ school, HolzingerSwineford1939)
coef(lm_sum)  # same as structural diffs
```

That coefficient representing the group difference looks familiar- it's the same value as we had for models `4a` and `4c`, where we looked at a group difference only for the visual factor.  As with those, this can be seen as a total effect of school on the visual 'factor'.



## Summary of differences

What should be apparent at this point is that the multigroup approach is an interaction of everything with the grouping variable.  Honestly I think this is rarely desirable theoretically, or at least we rarely do this in other modeling contexts.  Furthermore, group sizes may be prohibitively small given the number of parameters that need to be estimated.  In general, we may be interested in a specific group difference on some latent variable, possibly controlling for other effects in some fashion, and it is far simpler to specify such a model, and it is a notably simpler model as well.


## Measurement invariance

As a final note, in some cases we are instead looking for similarities across groups among the latent constructs, rather than differences.  This is especially the case in scale development, where one would like a measure to be consistent across groups of individuals (e.g. sex, age, race, etc.).  

Aside from general problems of 'accepting the null hypothesis', the basic idea is to test a restricted model (e.g. loadings, intercepts, etc. are equal) vs. the less restrictive one that assumes the differences across groups exist, and if the general fit of the models is not appreciably different, then one can claim equivalence across groups.  As a starting point, we assume <span class="emph">configural</span> equivalence, or in other words, that the factor structure is the same.  There is no point in testing measurement equivalence if there is not a similar factor structure.  The first more restricted model is that the loadings are equivalent.  The next is that observed variable intercepts are equivalent, followed by latent variable means, and finally residual variances/covariances.

I find in consulting and in published reports that researchers think that because they are interested in group differences that they are required to take a measurement invariance approach.  This is not the case at all, as our previous models have shown.

```{r measurement_invariance}
hs_model_4 <- ' 
  visual  =~ x1 + x2 + x3
  verbal  =~ x4 + x5 + x6
  speed   =~ x7 + x8 + x9 
'

semTools::measurementInvariance(
  model = hs_model_4, 
  data = HolzingerSwineford1939, 
  group = "school"
)


```

```{r measurement_invariance_current_approach}
# wow, from one line to this. this is an awful choice for an alternative
test.seq <- c("loadings","intercepts","means","residuals")
meq.list <- list()
for (i in 0:length(test.seq)) {
  if (i == 0L) {
    meq.label <- "configural"
    group.equal <- ""
    # long.equal <- ""
  } else {
    meq.label <- test.seq[i]
    group.equal <- test.seq[1:i]
    # long.equal <- test.seq[1:i]
  }
  meq.list[[meq.label]] <- semTools::measEq.syntax(
    configural.model = hs_model_baseline,
    data = HolzingerSwineford1939,
    ID.fac = "auto.fix.first",
    group = "school",
    group.equal = group.equal,
    # longFacNames = longFacNames,
    # long.equal = long.equal,
    return.fit = TRUE
  )
}

# and of course, this is still a borked function
semTools::compareFit(meq.list)


# means_only
test.seq <- c("means")
meq.list <- list()
for (i in 0:length(test.seq)) {
  if (i == 0L) {
    meq.label <- "configural"
    group.equal <- ""
    # long.equal <- ""
  } else {
    meq.label <- test.seq[i]
    group.equal <- test.seq[1:i]
    # long.equal <- test.seq[1:i]
  }
  meq.list[[meq.label]] <- semTools::measEq.syntax(
    configural.model = hs_model_baseline,
    data = HolzingerSwineford1939,
    ID.fac = "auto.fix.first",
    group = "school",
    group.equal = group.equal,
    # longFacNames = longFacNames,
    # long.equal = long.equal,
    return.fit = TRUE
  )
}

semTools::compareFit(meq.list)
```


[^fsr]: Note that lavaan allows one to take this two step approach while estimating proper standard errors given the measurement error associated with the latent variable scores.  The function is <span class="func">fsr</span>, but as of this writing, it is undergoing development, has been hidden from the user, and was not working.