---
title: "ICU Beds vs Covid-19 Cases"
author: by MC
output: 
  flexdashboard::flex_dashboard:
    theme: journal
    social: [ "twitter", "facebook", "linkedin" ] #, 'weibo', 'wechat'
    source_code: embed
    css: styles.css
---

```{r setup, include = FALSE}
library(flexdashboard)
library(tidyverse)
library(leaflet)
library(crosstalk)
library(htmltools)
```

Column {.tabset data-width=400}
-----------------------------------------------------------------------

```{r nytimes, eval = F}
us_counties0 <- readr::read_csv(
  "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv"
)

terr = c(
  'Puerto Rico',
  'Guam',
  'Northern Mariana Islands',
  'American Samoa'
)


county_trends = us_counties0 %>% 
  filter(!state %in% terr) %>% 
  arrange(state, county, date) %>% 
  group_by(state, county) %>% 
  mutate(
    daily_cases = cases - lag(cases, default = NA),
    daily_cases = if_else(is.na(daily_cases), cases, daily_cases),
    date_num = as.numeric(date)
  ) %>% 
  ungroup() %>% 
  left_join(tibble(state = state.name, state_abb = state.abb)) %>% 
  mutate(state_abb = as.factor(state_abb)) %>% 
  mutate_at(vars(county, state), tolower)

```

```{r prelim-county-data}

# source('_posts/2020-03-23-covid/get_county_data.R')
# 
# county_trends = get_county_data(parallel = TRUE)
# 
# # for testing
# save(county_trends, file = 'data/county_trends.RData')
load('data/county_trends.RData')

# remove clutter that isn't necessary for our purposes
county_trends = county_trends %>% 
  filter(!county %in% c('Unknown', 'Unassigned', 'Out-of-state')) %>% 
  mutate_at(vars(county, state), tolower)

# some counties will not have a hospital, many counties do not have a current
# count, some counties from covid data don't have a fips, so this match could be better
# county_trends = county_trends %>% 
#   group_by(state, county) %>% 
#   mutate(fips = ifelse(nchar(fips) == 4, paste0('0', fips), fips)) %>% 
#   fill(fips, .direction = 'downup') 

# county_current_counts = county_trends %>% 
#   group_by(county) %>% 
#   filter(date == max(date)) %>% 
#   ungroup() %>% 
#   mutate_at(vars(county, state), tolower)


# we will define current as the number of new cases in the past two weeks;

library(lubridate)
county_current_counts = county_trends %>% 
  arrange(state, county, date) %>% 
  group_by(state, county) %>% 
  filter(date == max(date) | (date == (max(date) - weeks(2)) | date == min(date)))

# numerous issues with dplyr inability to handle basic logical operations. This is ugly but appears to work.
county_current_counts = county_current_counts %>% 
  group_by(state, county) %>% 
  # because group by case_when is so very, very hard
  summarise(
    # N = n(), # for testing
    confirmed_last_2wk = ifelse(
      length(confirmed) == 1,
      confirmed,
      confirmed[length(confirmed)] - confirmed[length(confirmed)-1]),
    confirmed = last(confirmed)
  ) %>% 
  ungroup() 

# county_current_counts %>% count(state, county, sort=T) 
```

```{r data-processing}
# hosp_data0 = read_csv('https://github.com/jsfenfen/covid_hospitals_demographics/raw/master/data/processed/hospital_data.csv') 
# 
# # get full state names
# states = tibble(state_name = state.name, state = state.abb)
# 
# hosp_data = hosp_data0 %>%
#   left_join(states) %>%
#   rename(
#     fips = statecountyfips,
#     county = county_name
#     ) %>%
#   mutate(
#     county = tolower(county),
#     state_name = tolower(state_name)
#   )
# 
# # for testing
# save(hosp_data, file = 'data/hosp_data.RData')

load('data/hosp_data.RData')

current_hospital = hosp_data %>% 
  left_join(county_current_counts, by = c('county', 'state_name' = 'state')) %>% 
  mutate(
    icu_ratio = confirmed_last_2wk/all_adult_icu_beds,
    icu_ratio = if_else(is.infinite(icu_ratio), NA_real_, icu_ratio),
    icu_bed_utilization = all_adult_icu_beds * all_adult_icu_utilization / 100,
    hospital_name = str_to_title(hospital_name),
    city = str_to_title(city),
    county = str_to_title(county),
    lab = glue::glue(
      'Hospital: {hospital_name}<br>County confirmed cases: {confirmed}<br>ICU beds: {all_adult_icu_beds}'
    ),
    lab = map(lab, HTML),
    size_for_leaflet = 1000 * log(icu_ratio + 1), ## because leaflet is so very stupid about point size
  )

# current_hospital_collapse = current_hospital %>% 
#   group_by(county, state, state_name) %>% 
#   summarise(
#     confirmed = max(confirmed), 
#     icu_beds = sum(all_adult_icu_beds, na.rm = T)
#   ) %>% 
#   ungroup() 



```


```{r create-shared-data}
sd = current_hospital %>% 
    select(
      hospital_name,
      lat,
      lng,
      city,
      county,
      state,
      confirmed,
      confirmed_last_2wk,
      all_adult_icu_beds,
      icu_bed_utilization,
      icu_ratio,
      interns_residents,
      payroll_employees,
      lab,
      size_for_leaflet
    ) %>% 
    mutate(
      has_icu = factor(if_else(all_adult_icu_beds > 0, 'Yes', 'No'))
    ) %>% 
    arrange(desc(confirmed))

shared_data <- SharedData$new(sd)
```



### Number of total critical care beds vs. current confirmed C19 cases

```{r nation-map}
library(leaflet)

# color palette
pal <-
  colorNumeric(
    scico::scico(100, begin = .2, end = 1, palette = 'lajolla'),
    0:max(current_hospital$icu_ratio, na.rm = T),
    na.color = visibly::col2hex('gray85')
  )



# current_hospital_test = current_hospital %>%
#   mutate(
#     lab = glue::glue(
#       'Hospital: {hospital_name}County confirmed cases: {confirmed}ICU beds: {all_adult_icu_beds}'
#     ),
#     lab = map(lab, HTML),
#     size_for_leaflet = 1000 * log(icu_ratio + 1), ## because leaflet is so very stupid about point size
#   )

shared_data %>% 
  leaflet() %>% 
  addTiles() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  setView(-83.7, 42.3, zoom = 6) %>% 
  addCircles(
    lat = ~ lat,
    lng = ~ lng,
    # radius =  ~ size_for_leaflet * 2.1,  # possibly easier to modify here.
    radius =  6250,  # size in meters.
    color = ~ pal(icu_ratio),
    fillOpacity = .8,
    stroke = FALSE,
    label =  ~lab,
    labelOptions = labelOptions(
      direction = "bottom",
      style = list(
        "color" = "gray25",
        "box-shadow" = "3px 3px rgba(0,0,0,0.25)",
        "font-size" = "14px",
        "border-color" = "rgba(0,0,0,0.5)"
      )
    )
  )

```

### Data

Select rows to highlight those points on the map.  For more details on this data, see [the repository](https://github.com/jsfenfen/covid_hospitals_demographics).


```{r county-data-table}
shared_data %>% 
  DT::datatable(
    filter = "top",  # allows filtering on each column
    extensions = c(
      # "Buttons",  # add download buttons, etc
      "Scroller"  # for scrolling down the rows rather than pagination
    ),
    rownames = FALSE,  # remove rownames
    
    # can rename cols here, but probably easier to do in creation of shared data
    colnames = c(
      'Hospital' = 'hospital_name',
      'City' = 'city',
      'County' = 'county',
      'State' = 'state',
      'Confirmed' = 'confirmed',
      'Confirmed last 2 weeks' = 'confirmed_last_2wk',
      'ICU Beds' = 'all_adult_icu_beds',
      'Has ICU' = 'has_icu',
      'ICU Utilization' = 'icu_bed_utilization',
      'ICU Ratio' = 'icu_ratio',
      'Residents/Interns' = 'interns_residents',
      'Total Employ.' = 'payroll_employees'#,
      # Confirmed = 8,
    ),
    style = "bootstrap",
    class = "compact",
    width = "75%",
    # height = "25%",
    options = list(
      dom = "Blrtip",  # specify content (search box, etc)
      pageLength = 20,
      deferRender = TRUE,
      scrollY = 300,
      scrollX = '100px',
      # scroller = TRUE,
      columnDefs = list(
        list(
          visible = FALSE,
          targets = c(1:2, 13:14)
        )
      )#,
      # buttons = list(
      #   I("colvis"),  # turn columns on and off
      #   "csv",  # download as .csv
      #   "excel"  # download as .xlsx
      # )
    ),
  ) %>% 
  DT::formatRound(~ `ICU Ratio` + `ICU Utilization`, digits = 1)
```

### Details

Some specific columns are:

- `Confirmed`: Total number of confirmed cases at the county level
- `Confirmed last 2 weeks`: Total new cases over the last two weeks
- `ICU Beds`: Number of beds with CMS codes 0800 00801 00802 00803 00804 00805 00806 00807 00808 00810 00820 00830 00850
- `ICU Ratio`: Number of confirmed cases in the last 2 weeks to ICU beds 
- `ICU Utilization`: Utilization rate, as a percent, is calculated for all_icu_beds and for subtotal_acute_beds. all adult icu_beds * all adult icu utilization / 100
- `Residents/Interns`: total number of residents/interns  based on CMS documentation
- `Total Employ`: Total employment

Column {data-width=450}
-----------------------------------------------------------------------




```{r filters}

bscols(
  filter_select(
    id = "hospital",
    label = "Hospital",
    sharedData = shared_data,
    group = ~hospital_name
  ),
  filter_checkbox(
    id = "has_icu",
    label = "Has ICU?",
    sharedData = shared_data,
    group = ~has_icu,
    columns = 2,
    inline = TRUE
  ),
  widths = c(4, 2)
)

bscols(
  filter_slider(
    id     = "covid_cases",
    label  = "Confirmed county-wide covid cases",
    sharedData = shared_data,
    column = ~confirmed,
    step   = 10,
    # max    = 1000,
    round  = TRUE,
    sep    = "",
    ticks  = FALSE
  ),
  widths = c(5)
)
```




### Something else?

Or a long sentence moving at a certain pace down the page aiming for the
bottom-if not the bottom of this page then some other page-where it can rest, or
stop for a moment to think out the questions raised by its own (temporary)
existence, which ends when the page is turned, or the sentence falls out of the
mind that holds it (temporarily) in some kind of embrace, not necessarily an
ardent one, but more perhaps the kind of embrace enjoyed (or endured), by a wife
who has just waked up and is on her way to the bathroom in the morning to wash
her hair, and is bumped into by her husband, who has been lounging at the
breakfast table reading the newspaper, and doesn't see her coming out of the
bedroom, but, when he bumps into her, or is bumped into by her, raises his hands
to embrace her lightly, transiently, because he knows that if he gives her a
real embrace so early in the morning, before she has properly shaken the dreams
out of her head, and got her duds on, she won't respond, and may even become
slightly angry, and say something wounding, and so the husband invests in this
embrace not so much physical or emotional pressure as he might, because he
doesn't want to waste anything-with this sort of feeling, then, the sentence
passes through the mind more or less, and there is another way of describing the
situation too, which is to say that the sentence crawls through the mind like
something someone says to you while you are listening very hard to the FM radio,
some rock group there, with its thrilling sound, and so, with your attention or
the major part of it at least already rewarded, there is not much mind room you
can give to the remark, especially considering that you have probably just
quarreled with that person, the maker of the remark, over the radio being too
loud, or something like that, and the view you take, of the remark, is that
you'd really rather not hear it, but if you have to hear it, you want to listen
to it for the smallest possible length of time, and during a commercial, because
immediately after the commercial they're going to play a new rock song by your
favorite group, a cut that has never been aired before, and you want to hear it
and respond to it in a new way, a way that accords with whatever you're feeling
at the moment, or might feel, if the threat of new experience could be
(temporarily) overbalanced by the promise of possible positive benefits, or what
the mind construes as such, remembering that these are often, really, disguised
defeats (not that such defeats are not, at times, good for your character,
teaching you that it is not by success alone that one surmounts life, but that
setbacks, too, contribute to that roughening of the personality that, by
providing a textured surface to place against that of life, enables you to leave
slight traces, or smudges, on the face of human history-your mark) and after
all, benefit-seeking always has something of the smell of raw vanity about it,
as if you wished to decorate your own brow with laurel, or wear your medals to a
cookout, when the invitation had said nothing about them, and although the ego
is always hungry (we are told) it is well to remember that ongoing success is
nearly as meaningless as ongoing lack of success, which can make you sick, and
that it is good to leave a few crumbs on the table for the rest of your
brethren, not to sweep it all into the little beaded purse of your soul but to
allow others, too, part of the gratification, and if you share in this way you
will find the clouds smiling on you, and the postman bringing you letters, and
bicycles available when you want to rent them, and many other signs, however
guarded and limited, of the community's (temporary) approval of you, or at least
of it's willingness to let you believe (temporarily) that it finds you not so
lacking in commendable virtues as it had previously allowed you to think, from
its scorn of your merits, as it might be put, or anyway its consistent refusal
to recognize your basic humanness and its secret blackball of the project of
your remaining alive, made in executive session by its ruling bodies, which, as
everyone knows, carry out concealed programs of reward and punishment, under the
rose, causing faint alterations of the status quo, behind your back, at various
points along the periphery of community life, together with other enterprises
not dissimilar in tone, such as producing films that have special qualities, or
attributes, such as a film where the second half of it is a holy mystery, and
girls and women are not permitted to see it, or writing novels in which the
final chapter is a plastic bag filled with water, which you can touch, but not
drink: in this way, or ways, the underground mental life of the collectivity is
botched, or denied, or turned into something else never imagined by the
planners, who, returning from the latest seminar in crisis management and being
asked what they have learned, say they have learned how to throw up their hands;
the sentence meanwhile, although not insensible of these considerations, has a
festering conscience of its own, which persuades it to follow its star, and to
move with all deliberate speed from one place to another, without losing any of
the "riders" it may have picked up just being there, on the page, and turning
this way and that, to see what is over there, under that oddly-shaped tree, or
over there, reflected in the rain barrel of the imagination, even though it is
true that in our young manhood we were taught that short, punchy sentences were
best (but what did he mean? doesn't "punchy" mean punch-drunk? I think he
probably intended to say "short, punching sentences," meaning sentences that
lashed out at you, bloodying your brain if possible, and looking up the word
just now I came across the nearby "punkah," which is a large fan suspended from
the ceiling in India, operated by an attendant pulling a rope-that is what I
want for my sentence, to keep it cool!) we are mature enough now to stand the
shock of learning that much of what we were taught in our youth was wrong, or
improperly understood by those who were teaching it, or perhaps shaded a bit,
the shading resulting from the personal needs of the teachers, who as human
beings had a tendency to introduce some of their heart's blood into their work,
and sometimes this may not have been of the first water, this heart's blood, and
even if they thought they were moving the "knowledge" out, as the Board of
Education had mandated, they could have noticed that their sentences weren't
having the knockdown power of the new weapons whose bullets tumble end-over-end
(but it is true that we didn't have these weapons at that time) and they might
have taken into account the fundamental dubiousness of their project (but all
the intelligently conceived projects have been eaten up already, like the moon
and the stars) leaving us, in our best clothes, with only things to do like
conducting vigorous wars of attrition against our wives, who have now thoroughly
come awake, and slipped into their striped bells, and pulled sweaters over their
torsi, and adamantly refused to wear any bras under the sweaters, carefully
explaining the political significance of this refusal to anyone who will listen,
or look, but not touch, because that has nothing to do with it, so they say;
leaving us, as it were, with only things to do like floating sheets of Reynolds
Wrap around the room, trying to find out how many we can keep in the air at the
same time, which at least gives us a sense of participation, as though we were
Buddha, looking down at the mystery of your smile, which needs to be
investigated, and I think I'll do that right now, while there's still enough
light, if you'll sit down over there, in the best chair, and take off all your
clothes, and put your feet in that electric toe caddy (which prevents pneumonia)
and slip into this permanent press hospital gown, to cover your nakedness-why,
if you do all that, we'll be ready to begin! after I wash my hands, because you
pick up an amazing amount of exuviae in this city, just by walking around in the
open air, and nodding to acquaintances, and speaking to friends, and copulating
with lovers, in the ordinary course (and death to our enemies! by and by)-but
I'm getting a little uptight, just about washing my hands, because I can't find
the soap, which somebody has used and not put back in the soap dish, all of
which is extremely irritating, if you have a beautiful patient sitting in the
examining room, naked inside her gown, and peering at her moles in the mirror,
with her immense brown eyes following your every movement (when they are not
watching the moles, expecting them, as in a Disney nature film, to exfoliate)
and her immense brown head wondering what you're going to do to her, the pierced
places in the head letting that question leak out, while the therapist decides
just to wash his hands in plain water, and hang the soap! and does so, and then
looks around for a towel, but all the towels have been collected by the towel
service, and are not there, so he wipes his hands on his pants, in the back (so
as to avoid suspicious stains on the front) thinking: what must she think of me?
and, all this is very unprofessional and at-sea looking! trying to visualize the
contretemps from her point of view, if she has one (but how can she? she is not
in the washroom) and then stopping, because it is finally his own point of view
that he cares about and not hers, and with this firmly in mind, and a light,
confident step, such as you might find in the works of Bulwer-Lytton, he enters
the space she occupies so prettily and, taking her by the hand, proceeds to tear
off the stiff white hospital gown (but no, we cannot have that kind of
pornographic merde in this majestic and high-minded sentence, which will
probably end up in the Library of Congress) (that was just something that took
place inside his consciousness, as he looked at her, and since we know that
consciousness is always consciousness of something, she is not entirely without
responsibility in the matter) so, then, taking her by the hand, he falls into
the stupendous white puree of her abyss, no, I mean rather that he asks her how
long it has been since her last visit, and she says a fortnight, and he
shudders, and tells her that with a condition like hers (she is an immensely
popular soldier, and her troops win all their battles by pretending to be
forests, the enemy discovering, at the last moment, that those trees they have
eaten their lunch under have eyes and swords) (which reminds me of the
performance, in 1845, of Robert-Houdin, called The Fantastic Orange Tree,
wherein Robert-Houdin borrowed a lady's handkerchief, rubbed it between his
hands and passed it into the center of an egg, after which he passed the egg
into the center of a lemon, after which he passed the lemon into the center of
an orange, then pressed the orange between his hands, making it smaller and
smaller, until only a powder remained, whereupon he asked for a small potted
orange tree and sprinkled the powder thereupon, upon which the tree burst into
blossom, the blossoms turning into oranges, the oranges turning into
butterflies, and the butterflies turning into beautiful young ladies, who then
married members of the audience), a condition so damaging to real-time social
intercourse of any kind, the best thing she can do is give up, and lay down her
arms, and he will lie down in them, and together they will permit themselves a
bit of the old slap and tickle, she wearing only her Mr. Christopher medal, on
its silver chain, and he (for such is the latitude granted the professional
classes) worrying about the sentence, about its thin wires of dramatic tension,
which have been omitted, about whether we should write down some natural events
occurring in the sky (birds, lightning bolts), and about a possible coup d'etat
within the sentence, whereby its chief verb would be-but at this moment a
messenger rushes into the sentence, bleeding from a hat of thorns he's wearing,
and cries out: "You don't know what you're doing! Stop making this sentence, and
begin instead to make Moholy-Nagy cocktails, for those are what we really need,
on the frontiers of bad behavior!" and then he falls to the floor, and a trap
door opens under him, and he falls through that, into a damp pit where a blue
narwhal waits, its horn poised (but maybe the weight of the messenger, falling
from such a height, will break off the horn)-thus, considering everything very
carefully, in the sweet light of the ceremonial axes, in the run-mad
skimble-skamble of information sickness, we must make a decision as to whether
we should proceed, or go back, in the latter case enjoying the pathos of
eradication, in which the former case reading an erotic advertisement which
begins, How to Make Your Mouth a Blowtorch of Excitement (but wouldn't that
overtax our mouthwashes?) attempting, during the pause, while our burned mouths
are being smeared with fat, to imagine a better sentence, worthier, more
meaningful, like those in the Declaration of Independence, or a bank statement
showing that you have seven thousand kroner more than you thought you had-a
statement summing up the unreasonable demands that you make on life, and one
that also asks the question, if you can imagine these demands, why are they not
routinely met, tall fool? but of course it is not that query that this infected
sentence has set out to answer (and hello! to our girl friend, Rosetta Stone,
who has stuck by us through thick and thin) but some other query that we shall
some day discover the nature of, and here comes Ludwig, the expert on sentence
construction we have borrowed from the Bauhaus, who will-"Guten Tag,
Ludwig!"-probably find a way to cure the sentence's sprawl, by using the
improved way of thinking developed in Weimer-"I am sorry to inform you that the
Bauhaus no longer exists, that all of the great masters who formerly thought
there are either dead or retired, and that I myself have been reduced to
constructing books on how to pass the examination for police sergeant"-and
Ludwig falls through the Tugendhat House into the history of man-made objects; a
disappointment, to be sure, but it reminds us that the sentence itself is a
man-made object, not the one we wanted of course, but still a construction of
man, a structure to be treasured for its weakness, as opposed to the strength of
stones
